---
title: "Data Splitting"
author: "IZ Raad"
date: "2024-03-04"
output: html_document
---

```{r, include=FALSE}
library(tidyverse)
library(kableExtra) 
```

# Data Splitting

We are working with the [OpenFEMA Dataset: Individuals and Households Program](https://www.fema.gov/openfema-data-page/individuals-and-households-program-valid-registrations-v1) dataset, which has 71 variables and 20,967,572 observations. If order to better work with this dataset, we will split it into three smaller datasets:

[Dataset 1](#1): Randomly Select 50,000 IA Applications per Year

[Dataset 2](#2): Select IA Applications with Disaster Numbers Matching 5 Selected Disasters

[Dataset 3](#3): Select IA applications originating from Fayetteville, NC

## Loading the Dataset

```{r, message=FALSE, warning=FALSE, output=FALSE, results='hide'}
fema_data <- read_csv(
  "C:\\Users\\isabe\\OneDrive\\Desktop\\Stor_320\\Data_Splitting\\IndividualsAndHouseholdsProgramValidRegistrations.csv")
```

```{r}
kbl(cbind(fema_data, fema_data)) %>%
  kable_paper() %>%
  scroll_box(width = "100%", height = "200px")
```

## Dataset 1: Randomly Select 50,000 IA Applications per Year {#1}

First, we need to see what format our declarationDate column is in.

```{r}
str(fema_data$declarationDate)
```

We see that the datatype is POSIXct, meaning that we can easily extract the year into a separate column.

```{r, warning=FALSE, message=FALSE}
fema_data$year <- as.numeric(format(fema_data$declarationDate, "%Y"))
str(fema_data$year)
```

We need to split out dataset into 22 datasets: one for each year covered in our original dataset. These will be easier to work with.

```{r}
data2002 <- filter(fema_data, year==2002)
data2003 <- filter(fema_data, year==2003)
data2004 <- filter(fema_data, year==2004)
data2005 <- filter(fema_data, year==2005)
data2006 <- filter(fema_data, year==2006)
data2007 <- filter(fema_data, year==2007)
data2008 <- filter(fema_data, year==2008)
data2009 <- filter(fema_data, year==2009)
data2010 <- filter(fema_data, year==2010)
data2011 <- filter(fema_data, year==2011)
data2012 <- filter(fema_data, year==2012)
data2013 <- filter(fema_data, year==2013)
data2014 <- filter(fema_data, year==2014)
data2015 <- filter(fema_data, year==2015)
data2016 <- filter(fema_data, year==2016)
data2017 <- filter(fema_data, year==2017)
data2018 <- filter(fema_data, year==2018)
data2019 <- filter(fema_data, year==2019)
data2020 <- filter(fema_data, year==2020)
data2021 <- filter(fema_data, year==2021)
data2022 <- filter(fema_data, year==2022)
data2023 <- filter(fema_data, year==2023)
```

Let's look at one dataset to ensure that it looks the way we anticipate.

```{r}
kbl(cbind(data2002, data2002)) %>%
  kable_paper() %>%
  scroll_box(width = "100%", height = "200px")
```

Now that we have our datasets split into each year, we may now randomly select 50,000 IA Applications from each year.

```{r}
set.seed(5) # Set seed to ensure reproducibility.

smpl50000 <- function(dataset){
  smplindex = sample(1:nrow(dataset), 50000, replace=FALSE)
  return(dataset[smplindex,])
}

sample2002 <- smpl50000(data2002)
sample2003 <- smpl50000(data2003)
sample2004 <- smpl50000(data2004)
sample2005 <- smpl50000(data2005)
sample2006 <- smpl50000(data2006)
sample2007 <- smpl50000(data2007)
sample2008 <- smpl50000(data2008)
sample2009 <- smpl50000(data2009)
sample2010 <- smpl50000(data2010)
sample2011 <- smpl50000(data2011)
sample2012 <- smpl50000(data2012)
sample2013 <- smpl50000(data2013)
sample2014 <- smpl50000(data2014)
sample2015 <- smpl50000(data2015)
sample2016 <- smpl50000(data2016)
sample2017 <- smpl50000(data2017)
sample2018 <- smpl50000(data2018)
sample2019 <- smpl50000(data2019)
sample2020 <- smpl50000(data2020)
sample2021 <- smpl50000(data2021)
sample2022 <- smpl50000(data2022)
sample2023 <- smpl50000(data2023)
```

Let's look at one dataset to ensure that it looks the way we anticipate.

```{r}
kbl(cbind(sample2002, sample2002)) %>%
  kable_paper() %>%
  scroll_box(width = "100%", height = "200px")
```

Now that we have 22 datasets of length 50,000, we may combine them all into one dataset

```{r}
data1 <- rbind(sample2002, sample2003, sample2004, sample2005, sample2006,
               sample2007, sample2008, sample2009, sample2010, sample2011,
               sample2012, sample2013, sample2014, sample2015, sample2016,
               sample2017, sample2018, sample2019, sample2020, sample2021,
               sample2022, sample2023)

kbl(cbind(data1, data1)) %>%
  kable_paper() %>%
  scroll_box(width = "100%", height = "200px")
```

The creation of dataset 1 is complete.

```{r, eval=FALSE, include=FALSE}
write.csv(data1,
          "C:\\Users\\isabe\\OneDrive\\Desktop\\Stor_320\\Data_Splitting\\data1.csv",
          row.names=FALSE)
```

## Dataset 2: Select IA Applications with Disaster Numbers Matching 5 Selected Disasters {#2}

Our five selected disasters are Hurricane Florence, Hurricane Katrina, Baja California Earthquake (2010), 2020 California wildfires, and the 2013 Moore Tornado. The disaster numbers corresponding with these are 4393/4394/4401, 1602/1603/1604/1605, 1911, 4558/4569, and 4117, in this order. These numbers were found through the [FEMA disaster declarations search tool](https://www.fema.gov/disaster/declarations).

Now that we have the disaster numbers associated with each disaster we are interested in, we must filter our original dataset.

```{r}
data2 <- filter(fema_data, disasterNumber %in%
                  c(4393,4394,4401,1602,1603,1604,1605,1911,4558,4569,4117))

kbl(cbind(data2, data2)) %>%
  kable_paper() %>%
  scroll_box(width = "100%", height = "200px")
```

The creation of dataset 2 is complete.

```{r, eval=FALSE, include=FALSE}
write.csv(data2,
          "C:\\Users\\isabe\\OneDrive\\Desktop\\Stor_320\\Data_Splitting\\data2.csv",
          row.names=FALSE)
```

## Dataset 3: Select IA applications originating from Fayetteville, NC {#3}

This dataset will require a filtering process similar to Dataset 2. Instead of choosing observations with specific disaster numbers, we will choose observations whose city is Fayetteville, NC.

```{r}
data3 <- filter(fema_data, damagedStateAbbreviation == 'NC' & damagedCity == 'FAYETTEVILLE')

kbl(cbind(data3, data3)) %>%
  kable_paper() %>%
  scroll_box(width = "100%", height = "200px")
```

The creation of dataset 3 is complete.

```{r, eval=FALSE, include=FALSE}
write.csv(data3,
          "C:\\Users\\isabe\\OneDrive\\Desktop\\Stor_320\\Data_Splitting\\data3.csv",
          row.names=FALSE)
```