---
title: "Data Cleaning"
author: "Izzy Raad"
date: "2024-03-04"
output: html_document
---

```{r, include=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
library(tidyr)
library(dplyr)
library(rvest)
library(curl)
```

```{r, include=FALSE, message=FALSE, warning=FALSE}
fema_data <- read_csv(
  "C:\\Users\\isabe\\OneDrive\\Desktop\\Stor_320\\Data_Splitting\\IndividualsAndHouseholdsProgramValidRegistrations.csv")
data1 <- read_csv(
  "C:\\Users\\isabe\\OneDrive\\Desktop\\Stor_320\\Data_Splitting\\data1.csv")
data2 <- read_csv(
  "C:\\Users\\isabe\\OneDrive\\Desktop\\Stor_320\\Data_Splitting\\data2.csv")
data3 <- read_csv(
  "C:\\Users\\isabe\\OneDrive\\Desktop\\Stor_320\\Data_Splitting\\data3.csv")
IAdata1 <- read_csv(
  "C:\\Users\\isabe\\OneDrive\\Desktop\\Stor_320\\Data_Splitting\\IAdata1.csv")
```

We are working with the [OpenFEMA Dataset: Individuals and Households Program](https://www.fema.gov/openfema-data-page/individuals-and-households-program-valid-registrations-v1) dataset, which has 71 variables and 20,967,572 observations. In order to better work with this dataset, we will split it into three smaller datasets:

[Dataset 1](#1): Randomly Select 50,000 IA Applications per Year

[Dataset 2](#2): Select IA Applications with Disaster Numbers Matching 5 Selected Disasters

[Dataset 3](#3): Select IA applications originating from Fayetteville, NC

After we split our FEMA dataset into three smaller datasets, we must [adjust monetary variables for inflation](#adjinf).

# Data Splitting

Before we split our FEMA data into three datasets, let's get a sense of what our original data looks like.

```{r}
kbl(fema_data[1:15,]) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "center") %>%
  scroll_box(height = "300px")
```

## Dataset 1: Randomly Select 50,000 IA Applications per Year {#1}

First, we need to see what format our declarationDate column is in.

```{r}
str(fema_data$declarationDate)
```

We see that the datatype is POSIXct, meaning that we can easily extract the year into a separate column.

```{r, warning=FALSE, message=FALSE}
fema_data$year <- as.numeric(format(fema_data$declarationDate, "%Y"))
str(fema_data$year)
```

Now we may define functions that randomly select 50,000 IA Applications from each year.

```{r}
set.seed(5) # Set seed to ensure reproducibility.

smpl50000 <- function(dataset){
  smplindex = sample(1:nrow(dataset), 50000, replace=FALSE)
  return(dataset[smplindex,])
}
```

```{r, include=FALSE}
newname = paste('sample', 2002, sep="")
filtered <- filter(fema_data, year==2002)
assign(newname, smpl50000(filtered))
```

```{r, eval=FALSE}
for (yr in 2002:2023){
  newname = paste('sample', yr, sep="")
  filtered <- filter(fema_data, year==yr)
  assign(newname, smpl50000(filtered))
}
```

Let's look at one dataset to ensure that it looks the way we anticipate.

```{r}
kbl(sample2002[1:15,]) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "center") %>%
  scroll_box(height = "300px")
```

Now that we have 22 datasets of length 50,000, we may combine them all into one dataset

```{r, eval=FALSE}
data1 <- rbind(sample2002, sample2003, sample2004, sample2005, sample2006,
               sample2007, sample2008, sample2009, sample2010, sample2011,
               sample2012, sample2013, sample2014, sample2015, sample2016,
               sample2017, sample2018, sample2019, sample2020, sample2021,
               sample2022, sample2023)
```

```{r}
kbl(data1[1:15,]) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "center") %>%
  scroll_box(height = "300px")
```

The creation of dataset 1 is complete.

```{r, eval=FALSE, include=FALSE}
write.csv(data1,
          "C:\\Users\\isabe\\OneDrive\\Desktop\\Stor_320\\Data_Splitting\\data1.csv",
          row.names=FALSE)
```

## Dataset 2: Select IA Applications with Disaster Numbers Matching 5 Selected Disasters {#2}

Our five selected disasters are Hurricane Florence, Hurricane Katrina, Baja California Earthquake (2010), 2020 California wildfires, and the 2013 Moore Tornado. The disaster numbers corresponding with these are 4393/4394/4401, 1602/1603/1604/1605, 1911, 4558/4569, and 4117, in this order. These numbers were found through the [FEMA disaster declarations search tool](https://www.fema.gov/disaster/declarations).

Now that we have the disaster numbers associated with each disaster we are interested in, we must filter our original dataset.

```{r, eval=FALSE}
data2 <- filter(fema_data, disasterNumber %in%
                  c(4393,4394,4401,1602,1603,1604,1605,1911,4558,4569,4117))
```

```{r}
kbl(data2[1:15,]) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "center") %>%
  scroll_box(height = "300px")
```

The creation of dataset 2 is complete.

```{r, eval=FALSE, include=FALSE}
write.csv(data2,
          "C:\\Users\\isabe\\OneDrive\\Desktop\\Stor_320\\Data_Splitting\\data2.csv",
          row.names=FALSE)
```

## Dataset 3: Select IA applications originating from Fayetteville, NC {#3}

This dataset will require a filtering process similar to Dataset 2. Instead of choosing observations with specific disaster numbers, we will choose observations whose city is Fayetteville, NC.

```{r, eval=FALSE}
data3 <- filter(fema_data, damagedStateAbbreviation == 'NC' & damagedCity == 'FAYETTEVILLE')
```

```{r}
kbl(data3[1:15,]) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "center") %>%
  scroll_box(height = "300px")
```

The creation of dataset 3 is complete.

```{r, eval=FALSE, include=FALSE}
write.csv(data3,
          "C:\\Users\\isabe\\OneDrive\\Desktop\\Stor_320\\Data_Splitting\\data3.csv",
          row.names=FALSE)
```

# Adjusting for Inflation {#adjinf}

First, let's web scrape yearly consumer price indexes (CPI) from [the Federal Reserve Bank of Minneapolis](https://www.minneapolisfed.org/about-us/monetary-policy/inflation-calculator/consumer-price-index-1913-).

```{r}
URL.CPI="https://www.minneapolisfed.org/about-us/monetary-policy/inflation-calculator/consumer-price-index-1913-"

CPIdf = URL.CPI %>% read_html() %>%
  html_table(fill=TRUE) %>% .[[1]] %>%
  select("year"="Year", "CPI" = "Annual Average CPI(-U)")

kbl(CPIdf) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "center") %>%
  scroll_box(height = "300px")
```

Next we will create a function to adjust for inflation. Our datasets end in 2023, we should output a dollar amount adjusted to 2023. To adjust for inflation, we use the following formula:
$$P_{1} = P_{0}(\frac{CPI_{1}}{CPI_{0}})$$

```{r}
CPI2023 = filter(CPIdf, year == 2023)[[1, 2]]

price2023 <- function(price, yr){
  CPI = filter(CPIdf, year == yr)[[1, 2]]
  price2023 = price*CPI2023/CPI
  return(price2023)
}
```

Monetary variables for our three datasets include haAmount, ihpAmount, fipAmount, onaAmount, rpfvl, ppfvl, floodDamageAmount, foundationDamageAmount, roofDamageAmount, rentalAssistanceAmount, repairAmount, replacementAmount, and personalPropertyAmount. Let's make a function that takes in a column and outputs an inflation adjusted column.

```{r}
monetary <- c('haAmount', 'ihpAmount', 'fipAmount', 'onaAmount', 'rpfvl', 'ppfvl',
               'floodDamageAmount', 'foundationDamageAmount', 'roofDamageAmount',
               'rentalAssistanceAmount', 'repairAmount', 'replacementAmount',
               'personalPropertyAmount')

IAdata <- function(data){
  idlist = data$id
  yearlist = data$year
  IAdata = data
  for (col in monetary){
    colmn = data[[col]]
    newlist = c()
    for (i in 1:nrow(data)){
      newlist = c(newlist, price2023(colmn[i], yearlist[i]))
    }
    newname = paste('IA', col, sep="")
    newdf = as.data.frame(newlist)
    colnames(newdf) <- c(newname)
    newdf$id <- idlist
    IAdata = inner_join(IAdata, newdf, by='id')
  }
  return(IAdata)
}
```

Let us now apply this function to our three datasets.

```{r, eval=FALSE}
IAdata1 <- IAdata(data1)
IAdata2 <- IAdata(data2)
IAdata3 <- IAdata(data3)
```

Now, we check one of these new datasets to ensure that it is in the proper format.

```{r}
kbl(IAdata1[1:15,]) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "center") %>%
  scroll_box(height = "300px")
```

We have now adjusted all monetary variables for inflation.

```{r, eval=FALSE, include=FALSE}
# Saving new datasets

write.csv(IAdata1,
          "C:\\Users\\isabe\\OneDrive\\Desktop\\Stor_320\\Data_Splitting\\IAdata1.csv",
          row.names=FALSE)
write.csv(IAdata2,
          "C:\\Users\\isabe\\OneDrive\\Desktop\\Stor_320\\Data_Splitting\\IAdata2.csv",
          row.names=FALSE)
write.csv(IAdata3,
          "C:\\Users\\isabe\\OneDrive\\Desktop\\Stor_320\\Data_Splitting\\IAdata3.csv",
          row.names=FALSE)
```


















